---
title: "Setting Up BLAST Databases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting Up BLAST Databases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette guides you through the process of setting up BLAST databases for use with the `yourpackage` R package. We'll cover how to download and set up the SwissProt, nr (non-redundant), and nt (nucleotide) databases, as well as how to configure your environment to use these databases efficiently.

## Prerequisites

Before you begin, ensure that you have:

1.  BLAST+ installed on your system
2.  Sufficient disk space for the databases (SwissProt is \~1GB, nr is \~200GB, nt is \~150GB)
3.  A stable internet connection for downloading the databases

## Downloading and Setting Up Databases

### 1. Load the package

First, load the `yourpackage`:

```{r setup, eval=FALSE}
library(yourpackage)
```

### 2. Set up a directory for your BLAST databases

Choose a directory to store your BLAST databases:

```{r, eval=FALSE}
blast_db_dir <- "/path/to/blast/databases"
```

### 3. Download and extract the databases

Use the `download_blastdb` function to download and extract each database:

#### SwissProt

```{r, eval=FALSE}
download_blastdb("swissprot", blast_db_dir)
```

#### nr (non-redundant protein sequences)

```{r, eval=FALSE}
download_blastdb("nr", blast_db_dir)
```

#### nt (nucleotide sequences)

```{r, eval=FALSE}
download_blastdb("nt", blast_db_dir)
```

Note: Downloading the nr and nt databases may take a considerable amount of time due to their size.

## Setting Up the BLASTDB Environment Variable

To allow BLAST to find your databases easily, set up the BLASTDB environment variable:

### For Unix-like systems (Linux, macOS)

Add this line to your `.bashrc`, `.bash_profile`, or equivalent:

```         
export BLASTDB="/path/to/blast/databases:$BLASTDB"
```

### For Windows

Set the environment variable using the System Properties dialog or use this command in PowerShell:

```         
[Environment]::SetEnvironmentVariable("BLASTDB", "C:\path\to\blast\databases;" + $env:BLASTDB, "User")
```

## Using the Databases in R

In your R scripts or interactive sessions, use the `check_set_blastdb` function from the `yourpackage` to ensure the BLASTDB variable is set correctly:

```{r, eval=FALSE}
check_set_blastdb("/path/to/blast/databases")

# Now you can use perform_blast with any of the databases
blastx_result <- perform_blast(
  sequences = your_sequences,
  db_path = "/path/to/blast/databases",
  db_name = "swissprot",
  blast_type = "blastx",
  e_value = 1e-5,
  max_hits = 5,
  threads = 4,
  output_base = "my_blast_results",
  taxids = 32443  # Example: Teleostei taxon ID
)
```

## Conclusion

You have now set up the BLAST databases and configured your environment to use them with the `yourpackage` R package. Remember to keep your databases updated regularly for the most current sequence information.

For more information on BLAST databases and their usage, refer to the [NCBI BLAST documentation](https://www.ncbi.nlm.nih.gov/books/NBK279668/).
