---
title: "Downloading BLAST Databases with funseqR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Downloading BLAST Databases with funseqR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `funseqR` package provides tools for **fun**ctional **seq**uence annotation in **R**, including the ability to download BLAST databases efficiently. This vignette demonstrates how to use the `dwnld_ncbi_db` function to generate download commands and scripts for BLAST databases.

## Installation

You can install the `funseqR` package from GitHub using the `devtools` package:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("pygmyperch/funseqR")
```

## Loading the Package

First, load the `funseqR` package:

```{r setup}
library(funseqR)
```

## Generating Download Commands

The `dwnld_ncbi_db` function allows you to generate download commands for BLAST databases. You can choose from different download methods (wget, curl, rsync, or lftp) and control parameters like the number of simultaneous connections per file and the number of parallel downloads.

### Example: Downloading the 'nr' Database

Let's generate download commands for the 'nr' (non-redundant protein sequences) database using the lftp method:

```{r}
nr_download <- dwnld_ncbi_db(
  db_type = "nr",
  method = "lftp",
  parallel_downloads = 4,
  pget_n = 8,
  output_script = "download_nr.sh"
)
```

This command generates a shell script named "download_nr.sh" that can be used to download the 'nr' database.

### Viewing Generated Commands

You can view the generated lftp command:

```{r}
cat(nr_download$download_commands$lftp)
```

### Metadata and File Information

The function also provides metadata about the database, and information about the files to be downloaded:

```{r}
# View metadata
print(nr_download$metadata)

# View the first few files
head(nr_download$files)
```

## Running the Download Script

To actually download the database, you need to make the generated shell script executable and then run it. In R, you can do this using the `system` function:

```{r, eval=FALSE}
# Make the script executable
system("chmod +x download_nr.sh")

# Run the script
system("./download_nr.sh")
```

Note: These commands are not actually run in this vignette as they would initiate a large download.

Alternatively, if you prefer to use bash directly in the terminal:

``` bash
chmod +x download_nr.sh
./download_nr.sh
```

Remember to ensure you have sufficient storage space before running the download script, as BLAST databases can be quite large.

```{r}
print(nr_download$metadata$total_size_gb)
```

## Using Other Download Methods

The `dwnld_ncbi_db` function supports other download methods as well. Here are examples for wget, curl, and rsync:

```{r}
# wget
wget_download <- dwnld_ncbi_db("nt", method = "wget", output_script = "download_nt_wget.sh")

# curl
curl_download <- dwnld_ncbi_db("swissprot", method = "curl", output_script = "download_swissprot_curl.sh")

# rsync
rsync_download <- dwnld_ncbi_db("nr", method = "rsync", output_script = "download_nr_rsync.sh")
```

Each of these commands generates a script for downloading the specified database using the chosen method. You then simply make the script executable and run it as in the above lfpt example.

